---
title: "Day 1, Lab 2"
output: html_notebook
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

theme_set(theme_minimal())
```

## Lab II: Population Projections and Stability

In this section, we will project a population using the fertility, mortality and population data from the dataset from the first section of this lab. We will then compare these results to the growth rate and population structure implied by stable population theory. 

We will use the same data from the earlier exercises. 

First, we'll open up the data that we used in Lab 1.

```{r}
d <- read_csv("../bfdw_projection_data.csv") %>%
  # let's quickly repeat a few calculations that we made in Lab 1
         # change the radix to 1
  mutate(nLx = nLx/10^5, 
         # make fertility rates per woman, instead of per 1000 women
         nFx = nFx/1000) %>%
  group_by(country) %>%
         # calculate the fraction of the population in each age group
  mutate(nCx = nKx/sum(nKx)) %>%
  ungroup()
```

### Leslie matrices

We need to create a Leslie matrix which contains the fertility and survivorship information in order to project a population forward. Elements in the top line of the Leslie matrix are equal to 

$$
_nL_0 \cdot \frac{1}{2}(_nF_x + {_nF_{x+n}}\cdot\frac{_nL_{x+n}}{_nL_x}) \cdot f_{fab}
$$

and the subdiagonals are equal to

$$
\frac{_nL_{x+n}}{_nL_x}.
$$

Below is a function that helps to create a Leslie matrix. It takes four inputs: 

- a vector of $_nL_x$ values
- a vector of $_nF_x$ values
- the number of age groups. By default this is 10, because we are considering the five year age groups from ages 0--50.
- the fraction female at birth, which is set to a default value of 0.4886. 

With these inputs, the function `leslie` creates a square matrix with dimension equal to `n_age_groups` and fills in the top row and subdiagonal according to the equations above. 

```{r}
## function to define Leslie matrix, based on nLx and nFx values
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=10,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```

We can use this function to create a Leslie matrix using the $_nL_x$ and $_nF_x$ values for Australia:

```{r}
LAA <- leslie(nLx = d$nLx[d$country=="Australia"], 
              nFx = d$nFx[d$country=="Australia"])
LAA
```

We could also use $_nL_x$ and $_nF_x$ values for any combination of the different countries, for example, $_nL_x$ values from Australia and $_nF_x$ from Senegal:

```{r}
LAS <- leslie(nLx = d$nLx[d$country=="Australia"], 
              nFx = d$nFx[d$country=="Senegal"])
LAS
```


### Project the population 

We can now use the Leslie matrices to project a population forward in time. Let's start with the Australia-only Leslie matrix and project Australia's 2015 population forward by 250 years. 
Note that because we are using five-year age groups, the population projection happens in five-year steps. 

```{r}
n_age_groups <- 10 # 0-50 in 5yr age groups
n_projections <- 50 # want to project forward 50*5 = 250 years
initial_pop <- d$nKx[d$country=="Australia"] # set initial population to be Australia's population in 2015
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- LAA%*%K[,i-1] 
}

head(K[,1:5])
```

Now we have a matrix of populations by age for Australia projected into the future. We can use these numbers to investigate:

- The total population over time
- The share of the population in each age group over time
- The population growth rate over time

First, let's get the matrix `K` in a form that's a bit easier to plot. This involves making `K` into a dataframe in 'long' format, so that every row refers to a different year and age combination. 

```{r}
# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K)
colnames(Kdf) <- seq(from = 2015, to = (2015+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 45, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age))
head(Klong)
```
We'd like to caclulate the projected annual growth rate, $R$, of the Australian population over time.

First, let's recall that the annual growth rate is given by the equation

$$
K(t+5) = K(t) \cdot e^{5R}
$$
So, solving for $R$, we have

$$
R = \frac{1}{5} \cdot \log \frac{K(t+5)}{K(t)}
$$
Now we'll create a data frame that has the total population by year, and we'll use those total population values to 
calculate the annual growth rate over each 5-year time step

```{r}
# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  # this applies the formula for R that we derived above
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))
head(tot_pop)
```

OK. We have the results of our projection in nicely formatted dataframes.
Let's plot the results and see what our projection tells us.

First, let's look at the total population over time:

```{r}
# plot total population over time 
ggplot(data = tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Total population over time")
```


Next, let's investigate the growth rate over time

```{r}
# plot growth rate over time
ggplot(data = tot_pop, aes(x = year, y = R, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Growth rate of population over time")
```




Now we can plot total population, growth rate, and proportion in each age group over time.  

```{r}

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = year, y = proportion, group = age, color = age)) +
  geom_point() + geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position='bottom')+
  ggtitle("Proportion of population in each age group over time")
```

```{r}
# summarize the code we developed above in a function
project_population <- function(n_age_groups = 10,
                               n_projections = 50,
                               # the starting population
                               initial_pop,
                               # the Leslie matrix
                               L) {
  
  # define population matrix K
  K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
  K[,1] <- initial_pop[1:n_age_groups]
  
  # do the projection!
  for(i in 2:(n_projections+1)){
    K[,i] <- L%*%K[,i-1] 
  }
  
  # get K in a form that's easier to ggplot
  # make into a dataframe, name columns as years and add an age column
  Kdf <- as.data.frame(K)
  colnames(Kdf) <- seq(from = 2015, to = (2015+n_projections*5), by = 5)
  Kdf <- cbind(age = seq(from = 0, to = 45, by = 5), Kdf)
  
  # get in long format and then add proportion of population in each age group
  Klong <- Kdf %>% 
    gather(year, population, -age) %>%
    group_by(year) %>%
    mutate(proportion = population/sum(population), 
           age = as.factor(age))
  
  # summarize the projection with total population and growth rate
  tot_pop <- Klong %>% 
    group_by(year) %>%
    summarise(pop = sum(population)) %>%
    # this applies the formula for R that we derived above
    mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

  return(list(projection = Klong,
              tot_pop = tot_pop))
}
```

Let's illustrate how this function works by quickly repeating the projection
we made for Australia.
We feed Australia's starting values and Australia's Leslie matrix into the
function, and the result is an list with two objects: `projection` and `tot_pop`.

```{r}
proj_AA <- project_population(initial_pop = d$nKx[d$country == 'Australia'],
                              L =  LAA)

names(proj_AA)
```

`projection` has the long-form projection results, with the projected number of
women in each age group at each time period:

```{r}
proj_AA$projection
```

and `tot_pop` summarizes the projection with total population size and growth rate:

```{r}
proj_AA$tot_pop
```

## Exercises

> **Q1a**. Pick another country and project the population forward 50 time steps, like
> we just did for Australia. Copy/adapt the code above to make a plot of the
> trajectory of the total population size, growth rate and the age distributions for your projection.

BEGIN SOLUTION:

Let's pick Kenya.

```{r}
# make Leslie matrix with Kenyan rates
LKA <- leslie(nLx = d$nLx[d$country=="Kenya"], 
              nFx = d$nFx[d$country=="Kenya"])
# project using Kenyan matrix and current population
proj_KA <- project_population(initial_pop = d$nKx[d$country == 'Kenya'],
                              L =  LKA)

proj_KA$projection
proj_KA$tot_pop
```

Plot the total population size:
```{r}
# plot total population over time 
ggplot(data = proj_KA$tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Total population of Kenya over time")
```

the growth rate:
```{r}
# plot growth rate over time
ggplot(data = proj_KA$tot_pop, aes(x = year, y = R, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Growth rate of population of Kenya over time")
```

and age distribution:
```{r}
# plot proportion in each age group over time
ggplot(data = proj_KA$projection, aes(x = year, y = proportion, group = age, color = age)) +
  geom_point() + geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position='bottom')+
  ggtitle("Proportion of population of Kenya in each age group over time")
```

END SOLUTION

> **Q1b**. In lecture, we looked at an expression for the number of people in
> each age group of a stable population with growth rate $r$. We said
> $_nK_x(t) = B(t) e^{-rx} _nL_x/l_0$. See if you can convert this into
> an expression for the fraction of the population in each age group of a
> stable population, which we call $_nc_x$.

BEGIN SOLUTION:

By dividing both sides by the total population, $_{\infty}K_0(t)$,
the left hand side becomes $_nc_x$.
The first factor on the right-hand side is $B(t)/_{\infty}K_0(t)$,
which is the birth rate $b$. So we have

$$_nc_x(t) = b e^{-rx} {}_nL_x/l_0$$



END SOLUTION

> **Q1c**. See if you can estimate the intrinsic growth rate $r$ in the stable
> population by comparing the size of age groups in your projection. (NB: there is
> more than one way to do this.)

BEGIN SOLUTION:

We can find the intrinsic growth rate through the ratio of the proportion of age groups. 

In a stable population, the proportion of an age group $_nc_x$ can be expressed as: 

$$
_nc_x = be^{-rx}\frac{_nL_x}{l_0}
$$

where $b$ is the crude birth rate and $r$ is the intrinsic growth rate.

The ratio of the proportions of two age groups, for example, the age groups 0-5 and 5-10, can be thus expressed as:

$$
\frac{_5c_0}{_5c_5} = \frac{be^{-0r}\frac{_5L_0}{l_0}}{be^{-5r}\frac{_5L_5}{l_0}} = \frac{_5L_0}{e^{-5r}{_5L_5}}
$$

Solving for $r$:
$$
r = -\frac{1}{5}log(\frac{_5L_0}{\frac{_5c_0}{_5c_5}{_5L_5}}) 
$$
Values for $_nL_x$ are provided in the data. Values for $_nc_x$ can be extracted from the projected population after it stabilizes over time

```{r}
# extract values of ncx from the projected population in 2265, long after it stabilizes
# extract values of nLx directly from the data
# for two age groups, 0-5 and 5-10
KA_stable <- proj_KA$projection %>% filter(year == 2265) %>% slice(1:2) %>% mutate(nLx = d$nLx[d$country=="Kenya"][1:2]) 
KA_stable
```

```{r}
# calculate r
r <- -(1/5) * log(KA_stable$nLx[1]/(KA_stable$proportion[1]/KA_stable$proportion[2]*KA_stable$nLx[2]))
r

```

We see that this value of $r$ is the same as the value calculated during the projection. 
  
> **Q2a**. Now find a third country, and use the Leslie matrix from the country you
> picked for (1) to project the population of the country from (1) forward for 50
> time steps, but assume that the starting population is the population of the
> third country. Does this population converge to the same size? The same growth
> rate? The same age structure?

BEGIN SOLUTION:

Let's pick Singapore.
```{r}
# project using Kenyan rates and Singaporean initial population
proj_SI <- project_population(initial_pop = d$nKx[d$country == 'Singapore'],
                              L =  LKA)
proj_SI$projection
proj_SI$tot_pop
```

Plot the total population size:
```{r}
# plot total population over time 
ggplot(data = proj_SI$tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Total population of Singapore over time")
```

growth rate:
```{r}
# plot growth rate over time
ggplot(data = proj_SI$tot_pop, aes(x = year, y = R, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Growth rate of population of Singapore over time")
```

and age distribution:
```{r}
# plot proportion in each age group over time
ggplot(data = proj_KA$projection, aes(x = year, y = proportion, group = age, color = age)) +
  geom_point() + geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position='bottom')+
  ggtitle("Proportion of population of Singapore in each age group over time")
```

By comparing the plots to those from Q1a, we see that 

1. the total population size are different
2. their trajectories are similar (both grow exponentially). 
3. the intrinsic growth rates and stable age distributions are exactly the same .

END SOLUTION

> **Q2b**. Look at the trajectory of the growth rate in
> the projection from Q1 and the projection from Q2a. Which projection took longer
> to converge to stability? What do you think might affect the amount of time it
> takes to converge to stability? If you have time, you can check some of your
> predictions by making projections using other combinations of Leslie matrices
> and starting values.
 
BEGIN SOLUTION:

The projection using the Singaporean initial population in Q2a took longer to converge to stability. The difference between the initial growth rate and the intrinsic growth rate implied by the Leslie matrix affect the amount of time it takes to converge: the farther away the initial growth rate is from the intrinsic growth rate, the longer it takes to converge to stability. The difference between the initial Kenyan growth rate (0.02724) and the intrinsic growth rate implied by the Kenyan Leslie matrix (0.02376) is smaller than that between the initial Singaporean growth rate (0.01409) and the intrinsic growth rate. 

END SOLUTION





