---
title: "Berkeley Workshop in Formal Demography Pre-workshop Exercise Solutions: Population Projection"
author: Ayesha S. Mahmud
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(dplyr.summarise.inform = FALSE)
require(dplyr)
require(ggplot2)
require(tidyverse)
```

## Solutions to exercises

```{r, echo = FALSE, include = FALSE}
# set up the example
#Leslie Matrix
A <- matrix(c(0, 5, 10, 0.3, 0, 0, 0, 0.5, 0.2 ), nr = 3, byrow = TRUE)

# set up the intial population vector
K0 <- matrix(c(100,250,50), ncol = 1)

years <- 20
K.projected <- matrix(0, nrow = nrow(A), ncol = years+1)
K.projected[, 1] <- K0

for (i in 1:years)
{
	K.projected[, i + 1] <- A %*% K.projected[,i]
}
```

### 1. How is the total population changing over time?

```{r}
# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K.projected)
colnames(Kdf) <- seq(1, to = years+1, by = 1)
Kdf <- cbind(stage = c("Eggs", "Juvenile", "Adult"), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -stage) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         stage = as.factor(stage))


# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>% ungroup()

# plot total population over time 
ggplot(data = tot_pop, aes(x = as.numeric(year), y = pop, group = 1)) +
  geom_point() + geom_line()+
  xlab("Year") + ylab("K(t)")+
  ggtitle("Total population over time")

```

Looks like the fish population is growing fast. Possibly exponentially?

### 2. What is the long-term equilibrium age structure and annual growth rate?

First, let's plot the proportion in each age group over time to see how the age structure is changing.
```{r}

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = as.numeric(year), y = proportion, group = stage, color = stage)) +
  geom_point() + geom_line()+ 
  xlab("Year") + ylab("K(t)")+
  ggtitle("Proportion of population in each age group over time")
```
 
Looks like the age-structure becomes stable after a few years! This is one of the key concepts of stable population theory - **fixed mortality and fertility rates produce a stable age structure over time.**

Now, let's look at the growth rate. If we assume exponential growth (more on that during the workshop):

\( \textbf{K(t+1)} = \textbf{K (t)} e^{r}\) where \(r\) is the growth rate from projecting the population forward one year.

```{r}
# let's plot using base R so we now how to do it with and without ggplot2

# plot r for each year
tot_pop <- tot_pop %>% arrange(as.numeric(year))
lambPop<-log(tot_pop$pop[2:(years+1)]/tot_pop$pop[1:years])
matplot(1:years, lambPop, type="b", pch=1, ylab = "r", xlab = "Time")
```

Looks like in the long-run, the population grows annually at a constant rate, \(r\)!

### 3. What happens to the age structure and growth rate if we start with a different intial population and age distribution?


Let's set up the experiment again, now with a different starting population.

```{r}
# set up the example
#Leslie Matrix
A <- matrix(c(0, 5, 10, 0.3, 0, 0, 0, 0.5, 0.2 ), nr = 3, byrow = TRUE)

# set up the intial population vector
#K0 <- matrix(c(100,250,50), ncol = 1)
K0 <- matrix(c(50,75,300), ncol = 1) # changing this vector

years <- 20
K.projected <- matrix(0, nrow = nrow(A), ncol = years+1)
K.projected[, 1] <- K0

for (i in 1:years)
{
	K.projected[, i + 1] <- A %*% K.projected[,i]
}
```

Now lets repeat what we did earlier.

```{r}
# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K.projected)
colnames(Kdf) <- seq(1, to = years+1, by = 1)
Kdf <- cbind(stage = c("Eggs", "Juvenile", "Adult"), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -stage) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         stage = as.factor(stage))


# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>% ungroup()

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = as.numeric(year), y = proportion, group = stage, color = stage)) +
  geom_point() + geom_line()+ 
  xlab("Year") + ylab("K(t)")+
  ggtitle("Proportion of population in each age group over time")

# plot r for each year
tot_pop <- tot_pop %>% arrange(as.numeric(year))
lambPop<-log(tot_pop$pop[2:(years+1)]/tot_pop$pop[1:years])
matplot(1:years, lambPop, type="b", pch=1, ylab = "r", xlab = "Time")
```

Wow! They look nearly identical. This is another key concept of stable population theory - **regardless of the characteristics of the starting population i.e. age structure and size, if constant mortality and fertility rates are applied for a long time, the population will eventually attain the characteristics (long-run \(r\) and stable age distribution) that are intrinsic to those rates.**

### 4. What happens to the age structure and growth rate if we multiply the 1st row by a constant like 2?

```{r}
# change A
#Leslie Matrix
A <- matrix(c(0, 5*2, 10*2, 0.3, 0, 0, 0, 0.5, 0.2 ), nr = 3, byrow = TRUE)


# set up the intial population vector
K0 <- matrix(c(100,250,50), ncol = 1)

years <- 20
K.projected <- matrix(0, nrow = nrow(A), ncol = years+1)
K.projected[, 1] <- K0

for (i in 1:years)
{
	K.projected[, i + 1] <- A %*% K.projected[,i]
}

# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K.projected)
colnames(Kdf) <- seq(1, to = years+1, by = 1)
Kdf <- cbind(stage = c("Eggs", "Juvenile", "Adult"), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -stage) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         stage = as.factor(stage))


# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>% ungroup()

# plot total population over time 
ggplot(data = tot_pop, aes(x = as.numeric(year), y = pop, group = 1)) +
  geom_point() + geom_line()+
  xlab("Year") + ylab("K(t)")+
  ggtitle("Total population over time")


# plot proportion in each age group over time
ggplot(data = Klong, aes(x = as.numeric(year), y = proportion, group = stage, color = stage)) +
  geom_point() + geom_line()+ 
  xlab("Year") + ylab("K(t)")+
  ggtitle("Proportion of population in each age group over time")

# plot r for each year
tot_pop <- tot_pop %>% arrange(as.numeric(year))
lambPop<-log(tot_pop$pop[2:(years+1)]/tot_pop$pop[1:years])
matplot(1:years, lambPop, type="b", pch=1, ylab = "r", xlab = "Time")
```

Looks like we have a higher proportion of eggs (i.e. a 'younger' population), and a higher long-term growth rate.

### 5. What happens to the age structure and growth rate if you change the survival probabilities? 

```{r}
# change A
#Leslie Matrix
A <- matrix(c(0, 5, 10, 0.5*(0.3), 0, 0, 0, 0.5*(0.5), 0.5*(0.2) ), nr = 3, byrow = TRUE)


# set up the intial population vector
K0 <- matrix(c(100,250,50), ncol = 1)

years <- 20
K.projected <- matrix(0, nrow = nrow(A), ncol = years+1)
K.projected[, 1] <- K0

for (i in 1:years)
{
	K.projected[, i + 1] <- A %*% K.projected[,i]
}

# get K in a form that's easier to ggplot
# make into a dataframe, name columns as years and add an age column
Kdf <- as.data.frame(K.projected)
colnames(Kdf) <- seq(1, to = years+1, by = 1)
Kdf <- cbind(stage = c("Eggs", "Juvenile", "Adult"), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -stage) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         stage = as.factor(stage))


# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>% ungroup()


# plot total population over time 
ggplot(data = tot_pop, aes(x = as.numeric(year), y = pop, group = 1)) +
  geom_point() + geom_line()+
  xlab("Year") + ylab("K(t)")+
  ggtitle("Total population over time")

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = as.numeric(year), y = proportion, group = stage, color = stage)) +
  geom_point() + geom_line()+ 
  xlab("Year") + ylab("K(t)")+
  ggtitle("Proportion of population in each age group over time")

# plot r for each year
tot_pop <- tot_pop %>% arrange(as.numeric(year))
lambPop<-log(tot_pop$pop[2:(years+1)]/tot_pop$pop[1:years])
matplot(1:years, lambPop, type="b", pch=1, ylab = "r", xlab = "Time")
```

Again, we see a younger population, but there's a dramatic drop in the growth rate when we reduce survivorship at all ages. You can further modify the Leslie matrix to answer questions such as: What happens when we only change survivorship from juveniles to adult? What about when we change just the survivorship from eggs to juveniles? Does the adult survival rate have a strong influence on the population's future? Which transition is more important to the future of the population? 